---
title: "Getting started with cdiscmask"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with cdiscmask}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(42)
```

`cdiscmask` anonymizes CDISC-standard datasets (SDTM, ADaM) while preserving
data types and keeping cross-domain transformations consistent. Every change is
recorded in a **masking key** so the original data can be recovered.

We use the [`safetyData`](https://cran.r-project.org/package=safetyData)
package throughout — it ships complete SDTM and ADaM datasets from a fictional
clinical trial.

```{r load}
library(cdiscmask)
library(safetyData)
```

## The masking key

All state lives in a single `cdiscmask_key` object. Creating one is the first
step in every workflow.

```{r key}
key <- new_mask_key()
key
```

Passing the **same key** to every masking call is what guarantees that a
subject mapped to `SUBJ-0001` in DM is also `SUBJ-0001` in AE and LB, and
that all their dates are shifted by exactly the same number of days.

## Masking a single domain

`mask_domain()` detects sensitive columns automatically and applies the right
strategy for each type:

| Column type | Detection rule | Strategy |
|---|---|---|
| Subject ID | `USUBJID`, `SUBJID` | Replaced with opaque pseudonym |
| Date / datetime | DTC/DTM suffix (SDTM) or `Date` class (ADaM) | Shifted by per-subject random offset |
| Study day | DY suffix | Shifted by the same offset (integer arithmetic) |
| Verbatim / term | TERM, TRT, NAM, TEXT suffix | Replaced with `[REDACTED]` |
| Everything else | — | Unchanged |

```{r mask-dm}
dm_masked <- mask_domain(sdtm_dm, key, domain = "DM")

dm_masked[1:4, c("USUBJID", "SUBJID", "RFSTDTC", "RFENDTC", "AGE", "SEX", "RACE")]
```

Notice that `AGE`, `SEX`, and `RACE` are untouched — only identifiable
information is transformed. Dates shift by a random number of days while
preserving the date format, and both `USUBJID` and `SUBJID` receive pseudonyms.

## Masking multiple domains with a shared key

The key point of the cross-domain design: calling `mask_domain()` with the
**same key** on additional domains reuses all existing pseudonyms and date offsets.

```{r mask-multi}
ae_masked <- mask_domain(sdtm_ae, key, domain = "AE")
lb_masked <- mask_domain(sdtm_lb, key, domain = "LB")
```

We can verify that the same subject has the same pseudonym in all three domains:

```{r cross-domain-check}
orig_subj <- "01-701-1015"
pseudonym <- key$id_map[orig_subj]

cat("Pseudonym assigned:", pseudonym, "\n")
cat("Present in DM:", pseudonym %in% dm_masked$USUBJID, "\n")
cat("Present in AE:", pseudonym %in% ae_masked$USUBJID, "\n")
cat("Present in LB:", pseudonym %in% lb_masked$USUBJID, "\n")
```

Date consistency can be confirmed by checking that the offset is identical
across domains:

```{r date-consistency}
# The date shift for this subject (in days)
offset <- key$date_offsets[orig_subj]
cat("Date offset for", orig_subj, ":", offset, "days\n\n")

# Original vs masked reference start date in DM
cat("DM RFSTDTC original:", sdtm_dm$RFSTDTC[sdtm_dm$USUBJID == orig_subj][1], "\n")
cat("DM RFSTDTC masked:  ", dm_masked$RFSTDTC[dm_masked$USUBJID == pseudonym][1], "\n\n")

# Original vs masked AE date — shifted by the same offset
ae_orig <- sdtm_ae[sdtm_ae$USUBJID == orig_subj, ]
ae_mask <- ae_masked[ae_masked$USUBJID == pseudonym, ]
cat("AE AESTDTC original:", ae_orig$AESTDTC[1], "\n")
cat("AE AESTDTC masked:  ", ae_mask$AESTDTC[1], "\n")
```

## The `mask_cdisc()` convenience wrapper

When you have a named list of domains, `mask_cdisc()` handles key creation and
loops over all domains in one call.

```{r mask-cdisc}
masked <- mask_cdisc(list(
  DM = sdtm_dm,
  AE = sdtm_ae,
  LB = sdtm_lb,
  VS = sdtm_vs
))

# The key is attached as an attribute
key2 <- attr(masked, "key")
key2
```

```{r mask-cdisc-inspect}
# AE adverse event term is redacted; other columns are preserved
masked$AE[1:3, c("USUBJID", "AETERM", "AESTDTC", "AESTDY", "AESEV", "AESER")]
```

## ADaM datasets

`mask_cdisc()` works identically for ADaM. `Date`-class columns (e.g. `TRTSDT`)
are detected automatically.

```{r mask-adam}
adam_masked <- mask_cdisc(list(
  ADSL = adam_adsl,
  ADAE = adam_adae
))

adam_masked$ADSL[1:3, c("USUBJID", "TRTSDT", "TRTEDT", "AGE", "SEX", "ARM")]
```

## Reversing anonymization

`unmask_domain()` restores the original data exactly. The `cdiscmask_cols`
attribute that `mask_domain()` attaches to each masked data frame tells
`unmask_domain()` which columns to restore and how.

```{r unmask}
dm_back <- unmask_domain(masked$DM, attr(masked, "key"))

# Every column is identical to the original
identical(dm_back$USUBJID,  sdtm_dm$USUBJID)
identical(dm_back$RFSTDTC,  sdtm_dm$RFSTDTC)
identical(dm_back$AGE,      sdtm_dm$AGE)
```

```{r unmask-ae}
ae_back <- unmask_domain(masked$AE, attr(masked, "key"))

identical(ae_back$AETERM,   sdtm_ae$AETERM)
identical(ae_back$AESTDTC,  sdtm_ae$AESTDTC)
identical(ae_back$AESTDY,   sdtm_ae$AESTDY)
```

## Persisting the key

To share masked datasets with a collaborator and later restore the originals,
save the key separately from the data.

```{r save-key, eval = FALSE}
save_mask_key(key2, "masking_key.rds")

# Later, or in a different session:
key_loaded <- load_mask_key("masking_key.rds")
dm_restored <- unmask_domain(masked$DM, key_loaded)
```

The key must be stored and transmitted securely — it is the only artefact that
links pseudonyms back to real subject identifiers.

## Overriding column detection

If the automatic detection misses a column or picks up one it shouldn't, pass
a custom `cols` list to `mask_domain()`:

```{r custom-cols, eval = FALSE}
my_cols <- detect_sensitive_cols(sdtm_ae)
my_cols$free_text <- c(my_cols$free_text, "AEOUT")  # also redact outcome text

ae_custom <- mask_domain(sdtm_ae, key, cols = my_cols)
```

## Summary

The core workflow in three steps:

```r
# 1. Create (or load) a key
key <- new_mask_key()

# 2. Mask — reuse the same key for every domain
masked <- mask_cdisc(list(DM = sdtm_dm, AE = sdtm_ae, LB = sdtm_lb), key = key)

# 3. Save the key securely so you can unmask later
save_mask_key(key, "masking_key.rds")
```
